---
title: "Getting started with funconstrain"
description: >
  Solve a test problem, choose its dimensions, browse the catalogue, and use a
  stored reference safely.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`funconstrain` collects the 35 nonlinear least-squares problems used by Moré,
Garbow, and Hillstrom to test optimization software. We will start by fitting
Rosenbrock with `stats::optim()`, then see how to choose dimensions and use the
stored reference values.

## Solve your first problem

`funconstrain_problem()` returns a starting point and the functions needed by
an optimizer:

```{r optimize-rosen}
library(funconstrain)

problem <- funconstrain_problem("rosen")
fit <- stats::optim(
  par = problem$x0,
  fn = problem$fn,
  gr = problem$gr,
  method = "BFGS"
)

final_objective <- problem$fn(fit$par)
gradient_inf_norm <- max(abs(problem$gr(fit$par)))

data.frame(
  convergence = fit$convergence,
  final_objective = signif(final_objective, 3),
  gradient_inf_norm = signif(gradient_inf_norm, 3)
)
```

A convergence code of zero means that `stats::optim()` reports successful
completion. The small objective and gradient norm provide a second check on
the returned point.

The three fields used by `optim()` have direct roles: `x0` is the starting
vector, `fn` is the objective, and `gr` is its gradient.

```{r optimize-rosen-check, include=FALSE}
stopifnot(
  identical(fit$convergence, 0L),
  is.finite(final_objective),
  final_objective < 1e-12,
  is.finite(gradient_inf_norm),
  gradient_inf_norm < 1e-6
)
```

```{r resolved-contract, include=FALSE}
stopifnot(
  all(c("x0", "fn", "gr") %in% names(problem)),
  is.numeric(problem$x0),
  length(problem$x0) == 2L,
  is.function(problem$fn),
  is.function(problem$gr)
)
```

## Browse the catalogue

`funconstrain_catalog()` lists the 35 MGH problems in paper order. Here `n` is the
number of parameters and `m` is the number of residual functions:

```{r catalog}
catalog <- funconstrain_catalog()
catalog_view <- catalog[
  catalog$name %in% c("rosen", "jenn_samp", "ex_rosen", "chebyquad"),
]

knitr::kable(data.frame(
  Problem = paste0(catalog_view$title, " (", catalog_view$name, ")"),
  `Default n` = catalog_view$n_default,
  `Can choose n?` = ifelse(catalog_view$n_kind == "variable", "yes", "no"),
  `Default m` = catalog_view$m_default,
  `Can choose m?` = ifelse(catalog_view$m_configurable, "yes", "no"),
  row.names = NULL,
  check.names = FALSE
))
```

The [mathematical definitions](problem-definitions.html) give the exact
supported dimensions, standard start, and equations for every catalogue entry.

## Choose `n` or `m`

For a variable-dimension problem, supply `n`. For a problem whose residual
count can be chosen independently, supply `m`:

```{r configurations}
extended <- funconstrain_problem("ex_rosen", n = 4)
sampled <- funconstrain_problem("jenn_samp", m = 20)

data.frame(
  problem = c(extended$problem$name, sampled$problem$name),
  supplied = c("n = 4", "m = 20"),
  n = c(
    extended$configuration$effective$n,
    sampled$configuration$effective$n
  ),
  m = c(
    extended$configuration$effective$m,
    sampled$configuration$effective$m
  )
)
```

Extended Rosenbrock derives `m = 4` from the chosen `n = 4`. The
Jennrich--Sampson problem keeps its fixed `n = 2` and uses the requested
`m = 20`.

```{r configuration-check, include=FALSE}
stopifnot(
  identical(extended$configuration$effective, list(n = 4L, m = 4L)),
  identical(sampled$configuration$effective, list(n = 2L, m = 20L))
)
```

## Check a reference before using it

Before comparing a result with a stored minimum, check that the reference
applies to the problem you resolved:

```{r objective-gap}
fmin_reference <- problem$reference$fmin
objective_gap <- if (identical(fmin_reference$status, "applicable")) {
  final_objective - fmin_reference$value
} else {
  NA_real_
}

data.frame(
  reference_status = fmin_reference$status,
  objective_gap = signif(objective_gap, 3)
)
```

Minimum values and minimizer locations are checked separately. Extended
Rosenbrock at `n = 4` shows why:

```{r split-reference}
data.frame(
  reference = c("Minimum value", "Minimizer location"),
  status = c(
    extended$reference$fmin$status,
    extended$reference$xmin$status
  ),
  value_stored = c(
    !is.null(extended$reference$fmin$value),
    !is.null(extended$reference$xmin$value)
  )
)
```

The minimum value zero applies at `n = 4`, while the stored minimizer belongs
to the package's `n = 8` example. Check the status of the particular reference
you intend to use. The [`funconstrain_problem()`
reference](../reference/funconstrain_catalog.html) documents the complete
status and diagnostic vocabulary.

```{r reference-check, include=FALSE}
stopifnot(
  identical(fmin_reference$status, "applicable"),
  identical(extended$reference$fmin$status, "applicable"),
  identical(extended$reference$xmin$status, "not_applicable"),
  identical(extended$reference$xmin$source_configuration, list(n = 8L))
)
```

## Manufactured families

How much does a small objective value tell us about the solution? The answer depends on what we
want to recover. I added two problem families to make this distinction easier to explore:

- `psd_factorization()` looks for a low-rank factor of a positive semidefinite matrix. Errors in
  directions with small eigenvalues can contribute little to the objective, even when their relative
  error is large. Also, different factors can give exactly the same matrix, so comparing a factor
  directly with one stored answer would be misleading.
- `nonlinear_springs()` looks for the equilibrium of a spring network on a square grid. Here we can
  check both the values at grid points (the *field*) and their differences along edges. Small field
  error need not imply small edge error. The `gamma` control lets us move from a quadratic problem
  to a nonlinear one while keeping the solution and starting point fixed.

These are *manufactured* problems: we choose the solution first and construct a problem around it.
Each returns a minimizer for the chosen configuration, a starting point, and the usual objective and
derivatives. These aren't part of the numbered MGH catalogue, so use the constructors directly.

This small panel varies the positive target spectrum for factorization and pairs harmonic and
nonlinear springs using the same grid, target, and start:

```{r manufactured-panel}
panel <- list(
  factor_k10 = psd_factorization(kappa = 10, seed = 101),
  factor_k100 = psd_factorization(kappa = 100, seed = 101),
  springs_harmonic = nonlinear_springs(gamma = 0, seed = 101),
  springs_nonlinear = nonlinear_springs(gamma = 10, seed = 101)
)
```

Defaults give 64-by-4 factors (256 parameters) and an 8-by-8 interior spring grid (64 parameters).
Constructors restore `RNGkind()` and `.Random.seed`. This preserves stream continuation with R's
default Inversion normal generator, but not a cached Box–Muller draw. See the constructor help when
interleaving calls with a nondefault generator. Save the generated problem with `saveRDS()`
when a comparison needs the same arrays across sessions; seeds alone do not guarantee bitwise
agreement across numerical-library environments.

### Compare objective reduction and solution quality

Let's use BFGS again, then call `quality()` on each returned point. This example needs only
`funconstrain` and base R; run it after loading the package and creating `panel` above.

```{r manufactured-fit}
panel_results <- lapply(panel, function(problem) {
  initial_objective <- problem$fn(problem$x0)
  fit <- stats::optim(
    par = problem$x0,
    fn = problem$fn,
    gr = problem$gr,
    method = "BFGS",
    control = list(maxit = 1000, reltol = 1e-12)
  )
  quality <- problem$quality(fit$par)
  list(
    fit = fit,
    quality = quality,
    objective_ratio = problem$fn(fit$par) / initial_objective
  )
})

fit_summary <- do.call(
  rbind,
  lapply(names(panel_results), function(name) {
    result <- panel_results[[name]]
    data.frame(
      problem = name,
      convergence = result$fit$convergence,
      fn_calls = unname(result$fit$counts["function"]),
      gr_calls = unname(result$fit$counts["gradient"]),
      objective_ratio = signif(result$objective_ratio, 3)
    )
  })
)
fit_summary
```

`objective_ratio` compares the final objective with its value at the starting point. As before,
`convergence` is the optimizer's own assessment. The call counts come from `optim()` and exclude the
evaluations used for reporting. Objective and gradient calls can have different costs, so their sum
is not necessarily a useful measure of work.

Now for the solution errors:

```{r manufactured-quality}
quality_summary <- do.call(
  rbind,
  lapply(names(panel_results), function(name) {
    quality <- panel_results[[name]]$quality
    data.frame(
      problem = name,
      measure = names(quality),
      relative_error = signif(unlist(quality, use.names = FALSE), 3)
    )
  })
)
quality_summary
```

For factorization, `relative_product_error` measures the overall matrix error. It contains the same
information as the objective, on a relative scale. `max_relative_mode_error` checks each target
eigenvector separately, relative to its eigenvalue, and reports the largest error. This gives the
smaller eigenvalues a say in whether we consider the result accurate.

For springs, `relative_field_error` and `relative_edge_error` report the two errors separately.
All these measures compare with the known solution, rather than with the starting point. They
describe the parameters returned by the optimizer; the full results are still available in
`panel_results`, including each fit's `$par` and unrounded `$quality` values.

```{r manufactured-check, include=FALSE}
stopifnot(
  identical(panel$springs_harmonic$x0, panel$springs_nonlinear$x0),
  identical(panel$springs_harmonic$xmin, panel$springs_nonlinear$xmin),
  nrow(fit_summary) == length(panel),
  nrow(quality_summary) == 2L * length(panel),
  all(vapply(
    panel_results,
    function(result) {
      all(is.finite(unlist(result$quality))) &&
        is.finite(result$objective_ratio)
    },
    logical(1)
  ))
)
```

The [`psd_factorization()`](../reference/psd_factorization.html) and
[`nonlinear_springs()`](../reference/nonlinear_springs.html) reference pages define the controls and
quality measures. Both provide dense Hessians; storage grows quadratically in the parameter count.

## Next steps

For equations and supported dimensions, continue with [Mathematical
definitions of the test problems](problem-definitions.html). For optimizer
examples, see [Using funconstrain with mize](mize-integration.html) or [Using
the optimx bridge](optimx-bridge.html).
