---
title: "Getting started with resolved problems"
description: >
  Discover a test problem, resolve one exact configuration, run an optimizer,
  and interpret its reference metadata.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`funconstrain` provides the 35 nonlinear least-squares test problems from
Moré, Garbow, and Hillstrom. The resolver is the simplest entry point for
solver-neutral code: it turns a problem name and optional dimensions into one
concrete starting point, a set of dimension-checked callbacks, and metadata
describing the exact configuration.

## Find a problem

`funconstrain_catalog()` lists the available problems in paper order. Its
columns distinguish fixed and variable parameter dimensions (`n`) and show
whether the number of summand functions (`m`) can be configured.

```{r catalog}
library(funconstrain)

catalog <- funconstrain_catalog()
catalog[
  catalog$name %in% c("rosen", "jenn_samp", "ex_rosen", "chebyquad"),
  c(
    "number", "name", "n_kind", "n_default", "m_configurable", "m_default"
  )
]
```

Names are matched exactly. This makes a selected problem unambiguous in
scripts and saved experiment configurations.

## Resolve one exact problem

The Rosenbrock problem has fixed `n = 2` and `m = 2`, so no dimensions need to
be supplied:

```{r resolve-rosen}
rosen_problem <- funconstrain_problem("rosen")
names(rosen_problem)
rosen_problem$x0
rosen_problem$configuration
```

The resolved callbacks are:

- `fn(x)`, the scalar objective;
- `gr(x)`, its gradient;
- `he(x)`, its Hessian; and
- `fg(x)`, a combined objective-and-gradient calculation.

Unlike the raw factory interface, a resolved `x0` is always a numeric vector.
Each callback is pinned to the effective value of `n`, so an accidental vector
of another length fails at the problem boundary.

```{r resolved-contract, include=FALSE}
rosen_at_start <- rosen_problem$fg(rosen_problem$x0)
rosen_hessian <- rosen_problem$he(rosen_problem$x0)

stopifnot(
  identical(
    names(rosen_problem),
    c(
      "fn", "gr", "he", "fg", "x0", "problem", "configuration", "reference"
    )
  ),
  is.numeric(rosen_problem$x0),
  length(rosen_problem$x0) == 2L,
  length(rosen_at_start$fn) == 1L,
  length(rosen_at_start$gr) == 2L,
  identical(dim(rosen_hessian), c(2L, 2L))
)
```

## Run an optimizer

The callbacks can be passed directly to an optimizer. Here, `stats::optim()`
uses its BFGS method and the analytic gradient:

```{r optimize-rosen}
fit <- stats::optim(
  par = rosen_problem$x0,
  fn = rosen_problem$fn,
  gr = rosen_problem$gr,
  method = "BFGS"
)

final_objective <- rosen_problem$fn(fit$par)
gradient_inf_norm <- max(abs(rosen_problem$gr(fit$par)))

data.frame(
  convergence = fit$convergence,
  final_objective = signif(final_objective, 3),
  gradient_inf_norm = signif(gradient_inf_norm, 3)
)
```

A convergence code of zero means that `stats::optim()` reports successful
completion. The final objective and gradient infinity norm are evaluated
again through the resolved problem, independently of the optimizer's stored
objective value. Their small magnitudes provide a useful quality check for
this run; exact values and iteration counts can vary across R versions and
platforms.

```{r optimize-rosen-check, include=FALSE}
stopifnot(
  identical(fit$convergence, 0L),
  is.finite(final_objective),
  final_objective < 1e-12,
  is.finite(gradient_inf_norm),
  gradient_inf_norm < 1e-6
)
```

## Request dimensions explicitly

For a variable-dimension problem, pass `n`. For a problem with configurable
summand count, pass `m`:

```{r configurations}
extended <- funconstrain_problem("ex_rosen", n = 4)
sampled <- funconstrain_problem("jenn_samp", m = 20)

data.frame(
  problem = c(extended$problem$name, sampled$problem$name),
  requested = c("n = 4", "m = 20"),
  effective_n = c(
    extended$configuration$effective$n,
    sampled$configuration$effective$n
  ),
  effective_m = c(
    extended$configuration$effective$m,
    sampled$configuration$effective$m
  )
)
```

`configuration$requested` records normalized values supplied by the caller;
an omitted value remains `NULL`. `configuration$effective` records the
concrete integer values used to create the starting point and callbacks. Here,
Extended Rosenbrock derives `m = 4` from the requested `n = 4`, while Jennrich
and Sampson combines the requested `m = 20` with its default `n = 2`.

```{r configuration-check, include=FALSE}
stopifnot(
  identical(extended$configuration$requested, list(n = 4L, m = NULL)),
  identical(extended$configuration$effective, list(n = 4L, m = 4L)),
  identical(sampled$configuration$requested, list(n = NULL, m = 20L)),
  identical(sampled$configuration$effective, list(n = 2L, m = 20L))
)
```

## Decide whether a reference applies

Each resolved problem has separate `fmin` and `xmin` reference records. A
record contains a `value`, a `status`, a machine-readable `reason`, and
`source_configuration`, the encoded configuration rule associated with the
record. The status says whether a value is available and whether that rule
establishes applicability. The table below is built from four current problem
configurations.

```{r reference-data, include=FALSE}
reference_cases <- list(
  "rosen (defaults)" = rosen_problem,
  "ex_rosen (n = 4)" = extended,
  "jenn_samp (m = 20)" = sampled,
  "box_3d (defaults)" = funconstrain_problem("box_3d")
)

format_source_configuration <- function(configuration) {
  if (length(configuration) == 0L) {
    return("all supported configurations")
  }

  paste(
    names(configuration),
    unlist(configuration, use.names = FALSE),
    sep = " = ",
    collapse = ", "
  )
}

reference_table <- do.call(
  rbind,
  lapply(names(reference_cases), function(case_name) {
    records <- reference_cases[[case_name]]$reference
    data.frame(
      problem = case_name,
      reference = names(records),
      status = vapply(records, `[[`, character(1L), "status"),
      value_stored = ifelse(
        vapply(records, function(record) !is.null(record$value), logical(1L)),
        "yes",
        "no"
      ),
      source_configuration = vapply(
        records,
        function(record) {
          format_source_configuration(record$source_configuration)
        },
        character(1L)
      ),
      row.names = NULL
    )
  })
)
```

```{r reference-table, echo=FALSE}
knitr::kable(
  reference_table,
  col.names = c(
    "Problem configuration",
    "Reference",
    "Status",
    "Value stored",
    "Source configuration"
  )
)
```

Interpret the four possible statuses before using a stored value as a target:

- `applicable`: the value applies to the effective configuration;
- `not_applicable`: the stored value is retained for inspection, but its
  source configuration does not match this problem instance;
- `unavailable`: the package has no stored value, so `value` is `NULL`; and
- `unknown`: a value is stored, but the package has not encoded enough
  information to decide applicability.

No current catalogue entry returns `unknown`, but it remains part of the
public record vocabulary so consumers do not have to treat missing knowledge
as either applicable or inapplicable. Notice also that `fmin` and `xmin` can
have different statuses: at `ex_rosen(n = 4)`, the minimum value applies but
the stored minimizer is for `n = 8`. Always inspect the status of the specific
reference you intend to use.

```{r reference-check, include=FALSE}
stopifnot(
  identical(
    reference_table$status,
    c(
      "applicable", "applicable",
      "applicable", "not_applicable",
      "not_applicable", "not_applicable",
      "applicable", "unavailable"
    )
  ),
  identical(
    reference_cases[["box_3d (defaults)"]]$reference$xmin$value,
    NULL
  )
)

default_references <- lapply(
  catalog$name,
  function(name) funconstrain_problem(name)$reference
)
default_statuses <- unlist(lapply(
  default_references,
  function(records) vapply(records, `[[`, character(1L), "status")
))
stopifnot(!"unknown" %in% default_statuses)
```

## When to use a raw factory

Direct factories such as `rosen()` remain useful when you want to inspect the
lower-level problem definition or manage dimensions yourself. Their starting
point is a numeric vector for fixed-dimension problems and a function for
variable-dimension problems. The resolver is usually more convenient for
solver-neutral code because it returns one numeric start, pinned callbacks,
and explicit configuration and reference metadata.

See the [problem resolver reference](../reference/funconstrain_catalog.html)
for the complete return contract. To run the problems through the optional
`optimx` bridge, continue with [Using the optimx bridge](optimx-bridge.html).
