---
title: "Getting started with funconstrain"
description: >
  Solve a test problem, choose its dimensions, browse the catalogue, and use a
  stored reference safely.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`funconstrain` collects the 35 nonlinear least-squares problems used by Moré,
Garbow, and Hillstrom to test optimization software. We will start by fitting
Rosenbrock with `stats::optim()`, then see how to choose dimensions and use the
stored reference values.

## Solve your first problem

`funconstrain_problem()` returns a starting point and the functions needed by
an optimizer:

```{r optimize-rosen}
library(funconstrain)

problem <- funconstrain_problem("rosen")
fit <- stats::optim(
  par = problem$x0,
  fn = problem$fn,
  gr = problem$gr,
  method = "BFGS"
)

final_objective <- problem$fn(fit$par)
gradient_inf_norm <- max(abs(problem$gr(fit$par)))

data.frame(
  convergence = fit$convergence,
  final_objective = signif(final_objective, 3),
  gradient_inf_norm = signif(gradient_inf_norm, 3)
)
```

A convergence code of zero means that `stats::optim()` reports successful
completion. The small objective and gradient norm provide a second check on
the returned point.

The three fields used by `optim()` have direct roles: `x0` is the starting
vector, `fn` is the objective, and `gr` is its gradient.

```{r optimize-rosen-check, include=FALSE}
stopifnot(
  identical(fit$convergence, 0L),
  is.finite(final_objective),
  final_objective < 1e-12,
  is.finite(gradient_inf_norm),
  gradient_inf_norm < 1e-6
)
```

```{r resolved-contract, include=FALSE}
at_start <- problem$fg(problem$x0)
hessian <- problem$he(problem$x0)

stopifnot(
  identical(
    names(problem),
    c(
      "fn", "gr", "he", "fg", "x0", "problem", "configuration", "reference"
    )
  ),
  is.numeric(problem$x0),
  length(problem$x0) == 2L,
  length(at_start$fn) == 1L,
  length(at_start$gr) == 2L,
  identical(dim(hessian), c(2L, 2L))
)
```

## Browse the catalogue

`funconstrain_catalog()` lists all 35 problems in paper order. Here `n` is the
number of parameters and `m` is the number of residual functions:

```{r catalog}
catalog <- funconstrain_catalog()
catalog_view <- catalog[
  catalog$name %in% c("rosen", "jenn_samp", "ex_rosen", "chebyquad"),
]

knitr::kable(data.frame(
  number = catalog_view$number,
  name = catalog_view$name,
  title = catalog_view$title,
  n = ifelse(
    catalog_view$n_kind == "fixed",
    as.character(catalog_view$n_default),
    paste0("variable; default ", catalog_view$n_default)
  ),
  m = ifelse(
    catalog_view$m_configurable,
    paste0("choose; default ", catalog_view$m_default),
    paste0("fixed or follows n; default ", catalog_view$m_default)
  ),
  row.names = NULL
))
```

The [mathematical definitions](problem-definitions.html) give the supported
dimensions, standard start, and equations for every catalogue entry.

## Choose `n` or `m`

For a variable-dimension problem, supply `n`. For a problem whose residual
count can be chosen independently, supply `m`:

```{r configurations}
extended <- funconstrain_problem("ex_rosen", n = 4)
sampled <- funconstrain_problem("jenn_samp", m = 20)

data.frame(
  problem = c(extended$problem$name, sampled$problem$name),
  supplied = c("n = 4", "m = 20"),
  n = c(
    extended$configuration$effective$n,
    sampled$configuration$effective$n
  ),
  m = c(
    extended$configuration$effective$m,
    sampled$configuration$effective$m
  )
)
```

Extended Rosenbrock derives `m = 4` from the chosen `n = 4`. Jennrich and
Sampson keeps its fixed `n = 2` and uses the requested `m = 20`.

```{r configuration-check, include=FALSE}
stopifnot(
  identical(extended$configuration$requested, list(n = 4L, m = NULL)),
  identical(extended$configuration$effective, list(n = 4L, m = 4L)),
  identical(sampled$configuration$requested, list(n = NULL, m = 20L)),
  identical(sampled$configuration$effective, list(n = 2L, m = 20L))
)
```

## Check a reference before using it

Before comparing a result with a stored minimum, check that the reference
applies to the problem you resolved:

```{r objective-gap}
fmin_reference <- problem$reference$fmin
objective_gap <- if (identical(fmin_reference$status, "applicable")) {
  final_objective - fmin_reference$value
} else {
  NA_real_
}

data.frame(
  reference_status = fmin_reference$status,
  objective_gap = signif(objective_gap, 3)
)
```

Minimum values and minimizer locations are checked separately. Extended
Rosenbrock at `n = 4` shows why:

```{r split-reference}
data.frame(
  reference = c("Minimum value", "Minimizer location"),
  status = c(
    extended$reference$fmin$status,
    extended$reference$xmin$status
  ),
  value_stored = c(
    !is.null(extended$reference$fmin$value),
    !is.null(extended$reference$xmin$value)
  )
)
```

The minimum value zero applies at `n = 4`, while the stored minimizer belongs
to the package's `n = 8` example. Check the status of the particular reference
you intend to use. The [`funconstrain_problem()`
reference](../reference/funconstrain_problem.html) documents the complete
status and diagnostic vocabulary.

```{r reference-check, include=FALSE}
stopifnot(
  identical(fmin_reference$status, "applicable"),
  identical(extended$reference$fmin$status, "applicable"),
  identical(extended$reference$xmin$status, "not_applicable"),
  identical(extended$reference$xmin$source_configuration, list(n = 8L))
)

default_references <- lapply(
  catalog$name,
  function(name) funconstrain_problem(name)$reference
)
default_statuses <- unlist(lapply(
  default_references,
  function(records) vapply(records, `[[`, character(1L), "status")
))
stopifnot(!"unknown" %in% default_statuses)
```

## Next steps

For equations and supported dimensions, continue with [Mathematical
definitions of the test problems](problem-definitions.html). For optimizer
examples, see [Using funconstrain with mize](mize-integration.html) or [Using
the optimx bridge](optimx-bridge.html).
