---
title: "Using funconstrain with mize"
description: >
  Pass a test problem to mize, inspect the result, and repeat the setup over a
  small problem set.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
stopifnot(utils::packageVersion("mize") >= "0.2.5.9001")
```

`funconstrain` supplies the test problem; [`mize`](https://jlmelville.github.io/mize/)
supplies the optimizer. We will fit Rosenbrock with BFGS, look at the
optimization as it progresses, and then use a small harness to repeat the same
setup consistently.

> **Version requirement.** These examples use GitHub development `mize`
> 0.2.5.9001 or later. Install it with:
>
> ```r
> pak::pak("jlmelville/mize")
> ```

## Fit Rosenbrock with BFGS

Start by resolving Rosenbrock and passing its callbacks to `mize()`:

```{r direct-run}
library(funconstrain)

problem <- funconstrain_problem("rosen")
fg <- list(
  fn = problem$fn,
  gr = problem$gr,
  fg = problem$fg,
  hs = problem$he
)

fit <- mize::mize(
  problem$x0,
  fg,
  method = "BFGS",
  max_iter = 200,
  abs_tol = NULL,
  rel_tol = NULL,
  ginf_tol = 1e-6,
  step_tol = NULL,
  store_progress = TRUE
)

data.frame(
  status = fit$status,
  stopped_by = fit$terminate$what,
  iterations = fit$iter,
  objective_evaluations = fit$nf,
  gradient_evaluations = fit$ng,
  final_objective = signif(fit$f, 3),
  gradient_inf_norm = signif(fit$ginfn, 3)
)
```

`mize` reports that the gradient infinity-norm tolerance stopped this run. The
small final objective and gradient norm provide a useful check that the
returned point is good for this problem.

The callback list is a direct name-for-name match except for the Hessian:
`funconstrain` calls it `he`, while `mize` calls it `hs`.

```{r direct-run-check, include=FALSE}
stopifnot(
  identical(fit$status, "converged"),
  identical(fit$terminate$what, "ginf_tol"),
  is.finite(fit$f),
  fit$f < 1e-12,
  is.finite(fit$ginfn),
  fit$ginfn < 1e-6,
  is.data.frame(fit$progress),
  nrow(fit$progress) > 1L
)
```

## Follow the optimization

With `store_progress = TRUE`, `mize` records the initial state and then one row
per outer iteration. The objective trace shows the rapid early decrease and
the smaller improvements near the solution.

```{r objective-trace, fig.cap="BFGS objective trace for Rosenbrock. The initial objective is followed by the value at each outer iteration.", fig.alt="A line plot on a logarithmic vertical scale showing the Rosenbrock objective falling by many orders of magnitude over the BFGS iterations."}
objective_trace <- fit$progress$f
objective_trace[[1L]] <- problem$fn(problem$x0)

plot(
  seq_along(objective_trace) - 1L,
  objective_trace,
  type = "l",
  log = "y",
  xlab = "Outer iteration",
  ylab = "Objective"
)
```

The plot uses outer iteration to show the trajectory. The result table above
reports objective and gradient evaluations because methods can do different
amounts of work inside one iteration.

```{r objective-trace-check, include=FALSE}
stopifnot(
  length(objective_trace) == nrow(fit$progress),
  all(is.finite(objective_trace)),
  all(objective_trace > 0),
  tail(objective_trace, 1L) < objective_trace[[1L]]
)
```

## Repeat runs consistently

The direct call is all that is needed for one problem. Repeating it across
problems also calls for consistent stopping controls and result summaries.
`funconstrain` therefore installs a small sourceable example harness:

```{r load-harness}
harness_path <- system.file(
  "examples",
  "mize-harness.R",
  package = "funconstrain"
)
stopifnot(nzchar(harness_path))
source(harness_path)
```

Declare the problems and dimensions you want to run. Use `NA` to accept a
default dimension:

```{r suite}
manifest <- data.frame(
  name = c("rosen", "ex_rosen", "jenn_samp"),
  n = c(NA_integer_, 4L, NA_integer_),
  m = c(NA_integer_, NA_integer_, 20L)
)

suite <- run_mize_suite(
  manifest,
  method = "BFGS",
  controls = list(
    max_iter = 50,
    abs_tol = NULL,
    rel_tol = NULL,
    ginf_tol = 1e-6,
    step_tol = NULL
  )
)

suite_table <- do.call(rbind, lapply(suite$runs, function(run) {
  data.frame(
    problem = run$problem$name,
    n = run$problem$configuration$effective$n,
    m = run$problem$configuration$effective$m,
    status = run$outcome$status,
    objective_evaluations = run$work$native$nf,
    gradient_evaluations = run$work$native$ng,
    final_objective = signif(run$quality$final$objective, 3),
    gradient_inf_norm = signif(
      run$quality$final$gradient_inf_norm,
      3
    )
  )
}))
rownames(suite_table) <- NULL
suite_table
```

All three runs use the same method and stopping rule. The chosen rows exercise
a fixed problem, an explicit parameter dimension, and an explicit residual
count.

```{r suite-check, include=FALSE}
stopifnot(
  identical(names(suite$runs), manifest$name),
  all(vapply(
    suite$runs,
    function(run) identical(run$outcome$status, "converged"),
    logical(1L)
  )),
  identical(
    suite$runs$ex_rosen$problem$configuration$effective,
    list(n = 4L, m = 4L)
  ),
  identical(
    suite$runs$jenn_samp$problem$configuration$effective,
    list(n = 2L, m = 20L)
  )
)
```

The table uses the stopping status and evaluation totals reported by `mize`.
The final objective and gradient norm are checked afterward with the original
problem callbacks so every row receives the same final-quality calculation.

## Reproduce a run

Each harness result retains the supplied controls, problem dimensions, package
versions, platform, and development `mize` commit. A compact summary is often
enough for a report:

```{r provenance}
rosen_run <- suite$runs$rosen

data.frame(
  field = c(
    "Problem",
    "Dimensions",
    "Method",
    "funconstrain",
    "mize",
    "mize commit",
    "R platform"
  ),
  value = c(
    rosen_run$problem$title,
    paste0(
      "n = ", rosen_run$problem$configuration$effective$n,
      ", m = ", rosen_run$problem$configuration$effective$m
    ),
    rosen_run$method,
    rosen_run$provenance$funconstrain_version,
    rosen_run$provenance$mize_version,
    substr(rosen_run$provenance$mize_commit, 1L, 12L),
    rosen_run$provenance$platform
  ),
  row.names = NULL
)
```

For optimizer behavior and method selection, continue with the
[`mize` documentation](https://jlmelville.github.io/mize/). For problem
dimensions and stored references, see [Getting started with
funconstrain](getting-started.html) and [Mathematical definitions of the test
problems](problem-definitions.html).
