---
title: "Using funconstrain with mize"
description: >
  Pass a test problem to mize, inspect the result, and repeat the setup over a
  small problem set.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
stopifnot(utils::packageVersion("mize") >= "0.3.0")
```

`funconstrain` supplies the test problem; [`mize`](https://jlmelville.github.io/mize/)
supplies the optimizer. We will fit Rosenbrock with BFGS, look at the
optimization as it progresses, and then use a small harness to repeat the same
setup consistently.

> **Version requirement.** These examples use `mize` 0.3.0 or later. Install
> the current version with:
>
> ```r
> pak::pak("jlmelville/mize")
> ```

## Fit Rosenbrock with BFGS

Start by resolving Rosenbrock and passing its callbacks to `mize()`:

```{r direct-run}
library(funconstrain)

problem <- funconstrain_problem("rosen")
fg <- list(
  fn = problem$fn,
  gr = problem$gr,
  fg = problem$fg,
  hs = problem$he
)

fit <- mize::mize(
  problem$x0,
  fg,
  method = "BFGS",
  max_iter = 200,
  abs_tol = NULL,
  rel_tol = NULL,
  ginf_tol = 1e-6,
  step_tol = NULL,
  store_progress = TRUE
)

data.frame(
  status = fit$status,
  stopped_by = fit$terminate$what,
  iterations = fit$iter,
  objective_evaluations = fit$nf,
  gradient_evaluations = fit$ng,
  hessian_evaluations = fit$nh,
  inverse_hessian_evaluations = fit$nhi,
  final_objective = signif(fit$f, 3),
  gradient_inf_norm = signif(fit$ginfn, 3)
)
```

`mize` reports that the gradient infinity-norm tolerance stopped this run. The
small final objective and gradient norm provide a useful check that the
returned point is good for this problem.

The callback list is a direct name-for-name match except for the Hessian:
`funconstrain` calls it `he`, while `mize` calls it `hs`. BFGS does not use the
Hessian; `hs` is included so the same callback list can also be passed to
Hessian-based methods. Consequently `fit$nh` is zero. `fit$nhi` is also zero
because this adapter does not supply an inverse-Hessian callback; BFGS updates
its approximation internally rather than calling one.

```{r direct-run-check, include=FALSE}
stopifnot(
  identical(fit$status, "converged"),
  identical(fit$terminate$what, "ginf_tol"),
  is.finite(fit$f),
  fit$f < 1e-12,
  is.finite(fit$ginfn),
  fit$ginfn < 1e-6,
  identical(fit$nh, 0),
  identical(fit$nhi, 0),
  is.data.frame(fit$progress),
  nrow(fit$progress) > 1L
)
```

## Follow the optimization

With `store_progress = TRUE`, `mize` records the initial state and then one row
per outer iteration. The objective trace shows the rapid early decrease and
the smaller improvements near the solution. Progress storage does not request
an otherwise-unneeded objective value for the starting row, so the article
evaluates that value once for the plot. This display-only call is outside the
objective count in `fit$nf`.

```{r objective-trace, fig.cap="BFGS objective trace for Rosenbrock. The initial objective is followed by the value at each outer iteration.", fig.alt="A line plot on a logarithmic vertical scale showing the Rosenbrock objective falling by many orders of magnitude over the BFGS iterations."}
objective_trace <- fit$progress$f
objective_trace[[1L]] <- problem$fn(problem$x0)

plot(
  seq_along(objective_trace) - 1L,
  objective_trace,
  type = "l",
  log = "y",
  xlab = "Outer iteration",
  ylab = "Objective"
)
```

The plot uses outer iteration to show the trajectory. The result table above
reports objective, gradient, Hessian, and inverse-Hessian evaluations because
methods can do different amounts of work inside one iteration.

```{r objective-trace-check, include=FALSE}
stopifnot(
  length(objective_trace) == nrow(fit$progress),
  all(is.finite(objective_trace)),
  all(objective_trace > 0),
  tail(objective_trace, 1L) < objective_trace[[1L]]
)
```

## Repeat runs consistently

The direct call is all that is needed for one problem. Repeating it across
problems also calls for consistent stopping controls and result summaries.
`funconstrain` therefore installs a small sourceable example harness:

```{r load-harness}
harness_path <- system.file(
  "examples",
  "mize-harness.R",
  package = "funconstrain"
)
stopifnot(nzchar(harness_path))
source(harness_path)
```

Declare the problems and dimensions you want to run. Use `NA` to accept a
default dimension:

```{r suite}
problem_set <- data.frame(
  name = c("rosen", "ex_rosen", "jenn_samp"),
  n = c(NA_integer_, 4L, NA_integer_),
  m = c(NA_integer_, NA_integer_, 20L)
)

suite <- run_mize_suite(
  problem_set,
  method = "BFGS",
  controls = list(
    max_iter = 50,
    abs_tol = NULL,
    rel_tol = NULL,
    ginf_tol = 1e-6,
    step_tol = NULL
  )
)

suite_table <- do.call(rbind, lapply(suite$runs, function(run) {
  data.frame(
    problem = run$problem$name,
    n = run$problem$configuration$effective$n,
    m = run$problem$configuration$effective$m,
    status = run$outcome$status,
    objective_evaluations = run$work$native$nf,
    gradient_evaluations = run$work$native$ng,
    final_objective = signif(run$quality$final$objective, 3),
    gradient_inf_norm = signif(
      run$quality$final$gradient_inf_norm,
      3
    )
  )
}))
rownames(suite_table) <- NULL
suite_table
```

All three runs use the same method and stopping rule. The same harness handles
a fixed problem, a chosen parameter dimension, and a chosen residual count.
Objective values are problem-specific, so compare each run with its own start
or applicable reference rather than comparing final objectives across rows.

```{r suite-check, include=FALSE}
stopifnot(
  identical(names(suite$runs), problem_set$name),
  all(vapply(
    suite$runs,
    function(run) identical(run$outcome$status, "converged"),
    logical(1L)
  )),
  all(vapply(
    suite$runs,
    function(run) {
      identical(run$work$native$nh, 0) &&
        identical(run$work$native$nhi, 0)
    },
    logical(1L)
  )),
  identical(
    suite$runs$ex_rosen$problem$configuration$effective,
    list(n = 4L, m = 4L)
  ),
  identical(
    suite$runs$jenn_samp$problem$configuration$effective,
    list(n = 2L, m = 20L)
  )
)
```

The table uses the stopping status and evaluation totals reported by `mize`.
The harness also retains `nh` and `nhi` in `run$work$native`; both are zero for
these BFGS runs. The final objective and gradient norm are checked afterward
with the original problem callbacks, outside the optimizer's counts, so every
row receives the same final-quality calculation.

## Reproduce a run

Each harness result retains the supplied controls, problem dimensions, package
versions and platform. Installs from GitHub also retain the `mize` source
commit; release installs may not include commit metadata. A compact summary is
often enough for a report:

```{r provenance}
rosen_run <- suite$runs$rosen
mize_commit <- rosen_run$provenance$mize_commit
if (is.na(mize_commit)) {
  mize_commit <- "not recorded for this install"
} else {
  mize_commit <- substr(mize_commit, 1L, 12L)
}

data.frame(
  field = c(
    "Problem",
    "Dimensions",
    "Method",
    "R version",
    "funconstrain",
    "mize",
    "mize commit",
    "R platform"
  ),
  value = c(
    rosen_run$problem$title,
    paste0(
      "n = ", rosen_run$problem$configuration$effective$n,
      ", m = ", rosen_run$problem$configuration$effective$m
    ),
    rosen_run$method,
    rosen_run$provenance$r_version,
    rosen_run$provenance$funconstrain_version,
    rosen_run$provenance$mize_version,
    mize_commit,
    rosen_run$provenance$platform
  ),
  row.names = NULL
)
```

The result also retains the complete supplied control list in
`rosen_run$controls`.

For optimizer behavior and method selection, continue with the
[`mize` documentation](https://jlmelville.github.io/mize/). For problem
dimensions and stored references, see [Getting started with
funconstrain](getting-started.html) and [Mathematical definitions of the test
problems](problem-definitions.html).
