---
title: "Using funconstrain with mize"
description: >
  Resolve exact test problems, run development mize, and retain inspectable
  outcome, quality, work, and provenance records.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article shows a small, repeatable integration between `funconstrain` and
[`mize`](https://jlmelville.github.io/mize/). It resolves exact problem
instances, adapts their callback names, and keeps the optimizer's native
termination record separate from independent final-quality checks.

The first version supports GitHub development `mize` version 0.2.5.9001 or
later, rather than CRAN `mize` 0.2.5. Install the
supported version before running the examples:

```r
pak::pak("jlmelville/mize")
```

## Load the shared harness

The sourceable companion used here is installed with `funconstrain`. It remains
an unexported example rather than a package API:

```{r load-harness}
library(funconstrain)

harness_path <- system.file(
  "examples",
  "mize-harness.R",
  package = "funconstrain"
)
stopifnot(nzchar(harness_path))
source(harness_path)
```

At the callback boundary, `mize` uses `hs` for a Hessian while
`funconstrain` uses `he`. The adapter preserves the objective, gradient, and
combined callback and changes only that Hessian name:

```{r adapter}
problem <- funconstrain_problem("rosen")
names(as_mize_fg(problem))
```

The companion also wraps `fn`, `gr`, `fg`, and `hs` separately. These physical
call counts answer a different question from `mize`'s native `nf` and `ng`
evaluation totals, especially when one combined `fg` call supplies both an
objective and a gradient.

## Run one ordinary solve

`run_mize_problem()` owns the resolved starting point, callback list, and
method argument. Solver controls are supplied explicitly in a named list, so
the run record retains the choices that differ from `mize` defaults.

```{r ordinary-run}
ordinary <- run_mize_problem(
  "rosen",
  method = "BFGS",
  controls = list(
    max_iter = 200,
    abs_tol = NULL,
    rel_tol = NULL,
    ginf_tol = 1e-6,
    step_tol = NULL
  )
)

data.frame(
  status = ordinary$outcome$status,
  termination = ordinary$outcome$terminate$what,
  converged = ordinary$outcome$converged,
  final_objective = signif(ordinary$quality$final$objective, 3),
  gradient_inf_norm = signif(
    ordinary$quality$final$gradient_inf_norm,
    3
  )
)
```

`status`, `message`, `converged`, and `terminate` above are native `mize`
observations. The final objective and gradient infinity norm are evaluated
afterward through the original resolved callbacks. They are independent audit
observations, not a reinterpretation of the native termination status.

```{r ordinary-check, include=FALSE}
stopifnot(
  is.null(ordinary$conditions$error),
  length(ordinary$conditions$warnings) == 0L,
  identical(ordinary$outcome$status, "converged"),
  identical(ordinary$outcome$terminate$what, "ginf_tol"),
  is.character(ordinary$outcome$message),
  is.logical(ordinary$outcome$converged),
  identical(names(ordinary$outcome$terminate), c("what", "val")),
  is.finite(ordinary$quality$final$objective),
  is.finite(ordinary$quality$final$gradient_inf_norm),
  ordinary$quality$final$objective < ordinary$quality$initial$objective,
  ordinary$quality$final$gradient_inf_norm < 1e-6
)
```

The stored objective reference is kept as a structured record. Compare with
its value only when its status is `"applicable"`; a stored minimizer is not a
required equality target for a successful run.

```{r reference}
ordinary$reference[c("status", "value", "source_configuration")]
```

## Preserve a hard-budget outcome

A hard callback budget is a valid experimental outcome, not a malformed row.
This run permits only one combined objective-or-gradient evaluation:

```{r budget-run}
budget_limited <- run_mize_problem(
  "rosen",
  method = "BFGS",
  controls = list(
    max_iter = 3,
    max_fg = 1,
    abs_tol = NULL,
    rel_tol = NULL,
    grad_tol = NULL,
    ginf_tol = NULL,
    step_tol = NULL
  )
)

data.frame(
  status = budget_limited$outcome$status,
  termination = budget_limited$outcome$terminate$what,
  converged = budget_limited$outcome$converged,
  native_objective_available = !is.null(
    budget_limited$native_quality$objective
  ),
  audited_objective = signif(
    budget_limited$quality$final$objective,
    3
  )
)
```

Here `mize` legitimately omits its final objective because another optimizer
callback would violate the hard budget. The harness leaves that native field
absent. Its separately labelled audited objective is evaluated only after the
run, through the original callback and outside the optimizer callback counts.

```{r budget-check, include=FALSE}
stopifnot(
  identical(budget_limited$outcome$status, "budget_exhausted"),
  identical(budget_limited$outcome$converged, FALSE),
  identical(budget_limited$outcome$terminate$what, "max_fg"),
  is.null(budget_limited$native_quality$objective),
  is.finite(budget_limited$quality$final$objective),
  is.finite(budget_limited$quality$final$gradient_inf_norm)
)
```

## Inspect work and provenance

Keep native evaluation totals and physical callback invocations as two labelled
views rather than trying to reconcile them into one number:

```{r work}
ordinary$work
```

The record also keeps the exact configurations and environment provenance.
The commit identifies the development `mize` source used for this rendered
run; it records provenance without turning the article into a permanent commit
pin.

```{r provenance}
ordinary$problem
ordinary$provenance
```

## Run a small explicit manifest

For a small integration check, declare every problem instance. Use `NA` for a
default dimension and explicit values for variable `n` or configurable `m`:

```{r suite}
manifest <- data.frame(
  name = c("rosen", "ex_rosen", "jenn_samp"),
  n = c(NA_integer_, 4L, NA_integer_),
  m = c(NA_integer_, NA_integer_, 20L)
)

suite <- run_mize_suite(
  manifest,
  method = "BFGS",
  controls = list(max_iter = 10)
)

suite_table <- do.call(rbind, lapply(suite$runs, function(run) {
  data.frame(
    problem = run$problem$name,
    effective_n = run$problem$configuration$effective$n,
    effective_m = run$problem$configuration$effective$m,
    status = run$outcome$status,
    final_objective = signif(run$quality$final$objective, 3)
  )
}))
rownames(suite_table) <- NULL
suite_table
```

This is intentionally a small integration and reproducibility harness. It does
not rank methods or prescribe line searches and stopping controls. See the
[`mize` documentation](https://jlmelville.github.io/mize/) for optimizer
behavior and method selection; use this companion when exact `funconstrain`
problem resolution and an inspectable run record are the task.

```{r suite-check, include=FALSE}
stopifnot(
  identical(names(suite$runs), manifest$name),
  identical(
    suite$runs$ex_rosen$problem$configuration$effective,
    list(n = 4L, m = 4L)
  ),
  identical(
    suite$runs$jenn_samp$problem$configuration$effective,
    list(n = 2L, m = 20L)
  )
)
```
