---
title: "Mathematical definitions of the test problems"
description: >
  Residual equations, supported dimensions, and standard starting points for
  all 35 Moré--Garbow--Hillstrom problems in funconstrain.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

library(funconstrain)
catalog <- funconstrain_catalog()
stopifnot(
  nrow(catalog) == 35L,
  identical(catalog$number, seq_len(35L)),
  !anyDuplicated(catalog$name)
)
```

Use this page to find the residual equations, supported dimensions, and
standard starting point for any problem implemented by `funconstrain`. The
definitions follow Section 3 of
[Moré, Garbow, and Hillstrom
(1981)](https://doi.org/10.1145/355934.355936) using one notation throughout.
The authors' [Algorithm 566 source](https://netlib.org/toms/566.gz) is a
second primary source for resolving typographical and indexing ambiguities.

The formulas follow those sources; the dimensions and defaults describe the
current package. Where `funconstrain` supports a narrower range, that range is
shown. To run a problem, see
[Getting started with funconstrain](getting-started.html) or the
[`funconstrain_problem()` reference](../reference/funconstrain_catalog.html).

For $x=(x_1,\ldots,x_n)\in\mathbb{R}^n$, every problem below defines
$m$ residuals. The scalar callback returned by the package is

\[
F(x)=\sum_{i=1}^{m}r_i(x)^2,
\qquad \texttt{problem\$fn(x)}=F(x).
\]

Indices in the equations are one-based. A sum over an empty index range is
zero. $x^{(0)}$ denotes the standard starting point.

## Problem index

```{r problem-index, echo=FALSE}
problem_links <- sprintf(
  "[%s](#problem-%02d-%s)",
  catalog$title,
  catalog$number,
  gsub("_", "-", catalog$name, fixed = TRUE)
)
knitr::kable(
  data.frame(
    Number = catalog$number,
    Problem = problem_links,
    `Default n` = catalog$n_default,
    `Default m` = catalog$m_default,
    check.names = FALSE
  ),
  align = c("r", "l", "r", "r"),
  escape = FALSE
)
```

## 1. Rosenbrock Function {#problem-01-rosen}

**Dimensions:** $n=2$, $m=2$<br>
**Factory:** [`rosen()`](../reference/rosen.html)

\[
\begin{aligned}
r_1(x)&=10(x_2-x_1^2),\\
r_2(x)&=1-x_1.
\end{aligned}
\]

**Start:** $x^{(0)}=(-1.2,1)$

## 2. Freudenstein and Roth Function {#problem-02-freud-roth}

**Dimensions:** $n=2$, $m=2$<br>
**Factory:** [`freud_roth()`](../reference/freud_roth.html)

\[
\begin{aligned}
r_1(x)&=-13+x_1+\{(5-x_2)x_2-2\}x_2,\\
r_2(x)&=-29+x_1+\{(1+x_2)x_2-14\}x_2.
\end{aligned}
\]

**Start:** $x^{(0)}=(1/2,-2)$

## 3. Powell Badly Scaled Function {#problem-03-powell-bs}

**Dimensions:** $n=2$, $m=2$<br>
**Factory:** [`powell_bs()`](../reference/powell_bs.html)

\[
\begin{aligned}
r_1(x)&=10^4x_1x_2-1,\\
r_2(x)&=e^{-x_1}+e^{-x_2}-1.0001.
\end{aligned}
\]

**Start:** $x^{(0)}=(0,1)$

## 4. Brown Badly Scaled Function {#problem-04-brown-bs}

**Dimensions:** $n=2$, $m=3$<br>
**Factory:** [`brown_bs()`](../reference/brown_bs.html)

\[
\begin{aligned}
r_1(x)&=x_1-10^6,\\
r_2(x)&=x_2-2\times10^{-6},\\
r_3(x)&=x_1x_2-2.
\end{aligned}
\]

**Start:** $x^{(0)}=(1,1)$

## 5. Beale Function {#problem-05-beale}

**Dimensions:** $n=2$, $m=3$<br>
**Factory:** [`beale()`](../reference/beale.html)

For $i=1,2,3$, let $(c_1,c_2,c_3)=(1.5,2.25,2.625)$. Then

\[
r_i(x)=c_i-x_1(1-x_2^i).
\]

**Start:** $x^{(0)}=(1,1)$

## 6. Jennrich and Sampson Function {#problem-06-jenn-samp}

**Dimensions:** $n=2$, $m\ge2$; default $m=10$<br>
**Factory:** [`jenn_samp()`](../reference/jenn_samp.html)

For $i=1,\ldots,m$,

\[
r_i(x)=2+2i-e^{ix_1}-e^{ix_2}.
\]

**Start:** $x^{(0)}=(0.3,0.4)$

## 7. Helical Valley Function {#problem-07-helical}

**Dimensions:** $n=3$, $m=3$<br>
**Factory:** [`helical()`](../reference/helical.html)

Away from $(x_1,x_2)=(0,0)$, define

\[
\theta(x_1,x_2)=
\begin{cases}
\dfrac{1}{2\pi}\tan^{-1}(x_2/x_1),&x_1>0,\\[4pt]
\dfrac{1}{2\pi}\tan^{-1}(x_2/x_1)+\dfrac12,&x_1<0,\\[4pt]
\dfrac14\operatorname{sgn}(x_2),&x_1=0.
\end{cases}
\]

The residuals are

\[
\begin{aligned}
r_1(x)&=10\{x_3-10\theta(x_1,x_2)\},\\
r_2(x)&=10\{(x_1^2+x_2^2)^{1/2}-1\},\\
r_3(x)&=x_3.
\end{aligned}
\]

**Implementation note:** The package rejects the origin, where the objective
is undefined.

**Start:** $x^{(0)}=(-1,0,0)$

## 8. Bard Function {#problem-08-bard}

**Dimensions:** $n=3$, $m=15$<br>
**Factory:** [`bard()`](../reference/bard.html)

For $i=1,\ldots,15$, set $v_i=16-i$, $w_i=\min(i,v_i)$, and

\[
\begin{aligned}
(y_1,\ldots,y_{15})={}&(0.14,0.18,0.22,0.25,0.29,0.32,0.35,0.39,\\
&0.37,0.58,0.73,0.96,1.34,2.10,4.39).
\end{aligned}
\]

Then

\[
r_i(x)=y_i-\left(x_1+\frac{i}{v_ix_2+w_ix_3}\right).
\]

**Start:** $x^{(0)}=(1,1,1)$

## 9. Gaussian Function {#problem-09-gauss}

**Dimensions:** $n=3$, $m=15$<br>
**Factory:** [`gauss()`](../reference/gauss.html)

For $i=1,\ldots,15$, set $t_i=(8-i)/2$ and

\[
\begin{aligned}
(y_1,\ldots,y_{15})={}&(0.0009,0.0044,0.0175,0.0540,0.1295,\\
&0.2420,0.3521,0.3989,0.3521,0.2420,\\
&0.1295,0.0540,0.0175,0.0044,0.0009).
\end{aligned}
\]

Then

\[
r_i(x)=x_1\exp\left\{-\frac{x_2}{2}(t_i-x_3)^2\right\}-y_i.
\]

**Start:** $x^{(0)}=(0.4,1,0)$

## 10. Meyer Function {#problem-10-meyer}

**Dimensions:** $n=3$, $m=16$<br>
**Factory:** [`meyer()`](../reference/meyer.html)

For $i=1,\ldots,16$, set $t_i=45+5i$ and

\[
\begin{aligned}
(y_1,\ldots,y_{16})={}&(34780,28610,23650,19630,16370,13720,11540,9744,\\
&8261,7030,6005,5147,4427,3820,3307,2872).
\end{aligned}
\]

Then

\[
r_i(x)=x_1\exp\left(\frac{x_2}{t_i+x_3}\right)-y_i.
\]

**Start:** $x^{(0)}=(0.02,4000,250)$

## 11. Gulf Research and Development Function {#problem-11-gulf}

**Dimensions:** $n=3$, $3\le m\le100$; default $m=99$<br>
**Factory:** [`gulf()`](../reference/gulf.html)

For $i=1,\ldots,m$, define

\[
t_i=\frac{i}{100},\qquad
y_i=25+\{-50\log(t_i)\}^{2/3}.
\]

The residual is

\[
r_i(x)=\exp\left\{-\frac{|x_2-y_i|^{x_3}}{x_1}\right\}-t_i.
\]

**Source note:** The published equation has a typographical glyph where the
subtraction in $|x_2-y_i|$ should appear; the
corrected form above agrees with the authors' Algorithm 566 source and the
package.

**Start:** $x^{(0)}=(5,2.5,0.15)$

## 12. Box Three-Dimensional Function {#problem-12-box-3d}

**Dimensions:** $n=3$, $m\ge3$; default $m=20$<br>
**Factory:** [`box_3d()`](../reference/box_3d.html)

For $i=1,\ldots,m$, let $t_i=i/10$. Then

\[
r_i(x)=e^{-t_ix_1}-e^{-t_ix_2}-x_3(e^{-t_i}-e^{-i}).
\]

**Start:** $x^{(0)}=(0,10,20)$

## 13. Powell Singular Function {#problem-13-powell-s}

**Dimensions:** $n=4$, $m=4$<br>
**Factory:** [`powell_s()`](../reference/powell_s.html)

\[
\begin{aligned}
r_1(x)&=x_1+10x_2,\\
r_2(x)&=\sqrt5(x_3-x_4),\\
r_3(x)&=(x_2-2x_3)^2,\\
r_4(x)&=\sqrt{10}(x_1-x_4)^2.
\end{aligned}
\]

**Start:** $x^{(0)}=(3,-1,0,1)$

## 14. Wood Function {#problem-14-wood}

**Dimensions:** $n=4$, $m=6$<br>
**Factory:** [`wood()`](../reference/wood.html)

\[
\begin{aligned}
r_1(x)&=10(x_2-x_1^2),&r_2(x)&=1-x_1,\\
r_3(x)&=\sqrt{90}(x_4-x_3^2),&r_4(x)&=1-x_3,\\
r_5(x)&=\sqrt{10}(x_2+x_4-2),&
r_6(x)&=(x_2-x_4)/\sqrt{10}.
\end{aligned}
\]

**Start:** $x^{(0)}=(-3,-1,-3,-1)$

## 15. Kowalik and Osborne Function {#problem-15-kow-osb}

**Dimensions:** $n=4$, $m=11$<br>
**Factory:** [`kow_osb()`](../reference/kow_osb.html)

The data are

\[
\begin{aligned}
(y_1,\ldots,y_{11})={}&(0.1957,0.1947,0.1735,0.1600,0.0844,0.0627,\\
&0.0456,0.0342,0.0323,0.0235,0.0246),\\
(u_1,\ldots,u_{11})={}&(4,2,1,0.5,0.25,0.167,0.125,0.1,\\
&0.0833,0.0714,0.0625).
\end{aligned}
\]

For $i=1,\ldots,11$,

\[
r_i(x)=y_i-x_1\frac{u_i(u_i+x_2)}{u_i(u_i+x_3)+x_4}.
\]

**Start:** $x^{(0)}=(0.25,0.39,0.415,0.39)$

## 16. Brown and Dennis Function {#problem-16-brown-den}

**Dimensions:** $n=4$, $m\ge4$; default $m=20$<br>
**Factory:** [`brown_den()`](../reference/brown_den.html)

For $i=1,\ldots,m$, let $t_i=i/5$ and define

\[
\begin{aligned}
a_i(x)&=x_1+t_ix_2-e^{t_i},\\
b_i(x)&=x_3+x_4\sin(t_i)-\cos(t_i),\\
r_i(x)&=a_i(x)^2+b_i(x)^2.
\end{aligned}
\]

**Objective note:** This problem squares $a_i^2+b_i^2$ once more in the shared
objective $F(x)=\sum_i r_i(x)^2$.

**Start:** $x^{(0)}=(25,5,-5,1)$

## 17. Osborne 1 Function {#problem-17-osborne-1}

**Dimensions:** $n=5$, $m=33$<br>
**Factory:** [`osborne_1()`](../reference/osborne_1.html)

Let $t_i=10(i-1)$ and define the observations in consecutive blocks:

\[
\begin{aligned}
(y_1,\ldots,y_8)&=(0.844,0.908,0.932,0.936,0.925,0.908,0.881,0.850),\\
(y_9,\ldots,y_{16})&=(0.818,0.784,0.751,0.718,0.685,0.658,0.628,0.603),\\
(y_{17},\ldots,y_{24})&=(0.580,0.558,0.538,0.522,0.506,0.490,0.478,0.467),\\
(y_{25},\ldots,y_{32})&=(0.457,0.448,0.438,0.431,0.424,0.420,0.414,0.411),\\
y_{33}&=0.406.
\end{aligned}
\]

For $i=1,\ldots,33$,

\[
r_i(x)=y_i-\{x_1+x_2e^{-t_ix_4}+x_3e^{-t_ix_5}\}.
\]

**Start:** $x^{(0)}=(0.5,1.5,-1,0.01,0.02)$

## 18. Biggs EXP6 Function {#problem-18-biggs-exp6}

**Dimensions:** $n=6$, $m\ge6$; default $m=13$<br>
**Factory:** [`biggs_exp6()`](../reference/biggs_exp6.html)

For $i=1,\ldots,m$, let $t_i=i/10$ and

\[
y_i=e^{-t_i}-5e^{-10t_i}+3e^{-4t_i}.
\]

Then

\[
r_i(x)=x_3e^{-t_ix_1}-x_4e^{-t_ix_2}+x_6e^{-t_ix_5}-y_i.
\]

**Start:** $x^{(0)}=(1,2,1,1,1,1)$

## 19. Osborne 2 Function {#problem-19-osborne-2}

**Dimensions:** $n=11$, $m=65$<br>
**Factory:** [`osborne_2()`](../reference/osborne_2.html)

Let $t_i=(i-1)/10$. The observations, in consecutive blocks, are

\[
\begin{aligned}
(y_1,\ldots,y_8)&=(1.366,1.191,1.112,1.013,0.991,0.885,0.831,0.847),\\
(y_9,\ldots,y_{16})&=(0.786,0.725,0.746,0.679,0.608,0.655,0.616,0.606),\\
(y_{17},\ldots,y_{24})&=(0.602,0.626,0.651,0.724,0.649,0.649,0.694,0.644),\\
(y_{25},\ldots,y_{32})&=(0.624,0.661,0.612,0.558,0.533,0.495,0.500,0.423),\\
(y_{33},\ldots,y_{40})&=(0.395,0.375,0.372,0.391,0.396,0.405,0.428,0.429),\\
(y_{41},\ldots,y_{48})&=(0.523,0.562,0.607,0.653,0.672,0.708,0.633,0.668),\\
(y_{49},\ldots,y_{56})&=(0.645,0.632,0.591,0.559,0.597,0.625,0.739,0.710),\\
(y_{57},\ldots,y_{64})&=(0.729,0.720,0.636,0.581,0.428,0.292,0.162,0.098),\\
y_{65}&=0.054.
\end{aligned}
\]

For $i=1,\ldots,65$,

\[
\begin{aligned}
r_i(x)=y_i-&\{x_1e^{-t_ix_5}+x_2e^{-x_6(t_i-x_9)^2}\\
&+x_3e^{-x_7(t_i-x_{10})^2}+x_4e^{-x_8(t_i-x_{11})^2}\}.
\end{aligned}
\]

**Start:**
$x^{(0)}=(1.3,0.65,0.65,0.7,0.6,3,5,7,2,4.5,5.5)$

## 20. Watson Function {#problem-20-watson}

**Dimensions:** $2\le n\le31$, $m=31$; default $n=6$<br>
**Factory:** [`watson()`](../reference/watson.html)

For $i=1,\ldots,29$, define

\[
\begin{aligned}
t_i&=i/29,\\
a_i(x)&=\sum_{j=2}^{n}(j-1)x_jt_i^{j-2},\\
b_i(x)&=\sum_{j=1}^{n}x_jt_i^{j-1}.
\end{aligned}
\]

The residuals are

\[
\begin{aligned}
r_i(x)&=a_i(x)-b_i(x)^2-1,&&i=1,\ldots,29,\\
r_{30}(x)&=x_1,\\
r_{31}(x)&=x_2-x_1^2-1.
\end{aligned}
\]

**Start:** $x^{(0)}=(0,\ldots,0)$

## 21. Extended Rosenbrock Function {#problem-21-ex-rosen}

**Dimensions:** even $n\ge2$, $m=n$; default $n=8$<br>
**Factory:** [`ex_rosen()`](../reference/ex_rosen.html)

For $k=1,\ldots,n/2$,

\[
\begin{aligned}
r_{2k-1}(x)&=10(x_{2k}-x_{2k-1}^2),\\
r_{2k}(x)&=1-x_{2k-1}.
\end{aligned}
\]

**Start:** Repeat $(-1.2,1)$ exactly $n/2$ times.

## 22. Extended Powell Function {#problem-22-ex-powell}

**Dimensions:** $n\ge4$ and divisible by 4, $m=n$; default $n=20$<br>
**Factory:** [`ex_powell()`](../reference/ex_powell.html)

For $k=1,\ldots,n/4$, let $a=4k-3$. Then

\[
\begin{aligned}
r_a(x)&=x_a+10x_{a+1},\\
r_{a+1}(x)&=\sqrt5(x_{a+2}-x_{a+3}),\\
r_{a+2}(x)&=(x_{a+1}-2x_{a+2})^2,\\
r_{a+3}(x)&=\sqrt{10}(x_a-x_{a+3})^2.
\end{aligned}
\]

**Naming note:** MGH calls this the “Extended Powell singular” problem; the
package title is shortened.

**Start:** Repeat $(3,-1,0,1)$ exactly $n/4$ times.

## 23. Penalty Function I {#problem-23-penalty-1}

**Dimensions:** $n\ge1$, $m=n+1$; default $n=25$<br>
**Factory:** [`penalty_1()`](../reference/penalty_1.html)

With $a=10^{-5}$,

\[
\begin{aligned}
r_j(x)&=\sqrt{a}(x_j-1),&&j=1,\ldots,n,\\
r_{n+1}(x)&=\sum_{j=1}^{n}x_j^2-\frac14.
\end{aligned}
\]

**Start:** $x_j^{(0)}=j$

## 24. Penalty Function II {#problem-24-penalty-2}

**Dimensions:** $n\ge1$, $m=2n$; default $n=25$<br>
**Factory:** [`penalty_2()`](../reference/penalty_2.html)

With $a=10^{-5}$, the residuals are

\[
\begin{aligned}
r_1(x)&=x_1-0.2,\\
r_i(x)&=\sqrt a\{e^{x_i/10}+e^{x_{i-1}/10}-e^{i/10}-e^{(i-1)/10}\},
&&i=2,\ldots,n,\\
r_{n+i-1}(x)&=\sqrt a\{e^{x_i/10}-e^{-0.1}\},
&&i=2,\ldots,n,\\
r_{2n}(x)&=\sum_{j=1}^{n}(n-j+1)x_j^2-1.
\end{aligned}
\]

**Edge case:** For $n=1$, the two middle ranges are empty.

**Start:** $x_j^{(0)}=1/2$

## 25. Variably Dimensioned Function {#problem-25-var-dim}

**Dimensions:** $n\ge1$, $m=n+2$; default $n=30$<br>
**Factory:** [`var_dim()`](../reference/var_dim.html)

Define $s(x)=\sum_{j=1}^{n}j(x_j-1)$. Then

\[
\begin{aligned}
r_j(x)&=x_j-1,&&j=1,\ldots,n,\\
r_{n+1}(x)&=s(x),\\
r_{n+2}(x)&=s(x)^2.
\end{aligned}
\]

**Start:** $x_j^{(0)}=1-j/n$

## 26. Trigonometric Function {#problem-26-trigon}

**Dimensions:** $n\ge1$, $m=n$; default $n=30$<br>
**Factory:** [`trigon()`](../reference/trigon.html)

For $i=1,\ldots,n$,

\[
r_i(x)=n-\sum_{j=1}^{n}\cos(x_j)+i\{1-\cos(x_i)\}-\sin(x_i).
\]

**Start:** $x_j^{(0)}=1/n$

## 27. Brown Almost-Linear Function {#problem-27-brown-al}

**Dimensions:** $n\ge1$, $m=n$; default $n=30$<br>
**Factory:** [`brown_al()`](../reference/brown_al.html)

\[
\begin{aligned}
r_i(x)&=x_i+\sum_{j=1}^{n}x_j-(n+1),&&i=1,\ldots,n-1,\\
r_n(x)&=\prod_{j=1}^{n}x_j-1.
\end{aligned}
\]

**Edge case:** For $n=1$, the first range is empty.

**Start:** $x_j^{(0)}=1/2$

## 28. Discrete Boundary Value Function {#problem-28-disc-bv}

**Dimensions:** $n\ge1$, $m=n$; default $n=35$<br>
**Factory:** [`disc_bv()`](../reference/disc_bv.html)

Let $h=1/(n+1)$, $t_i=ih$, and set the boundary values
$x_0=x_{n+1}=0$. For $i=1,\ldots,n$,

\[
r_i(x)=2x_i-x_{i-1}-x_{i+1}+\frac{h^2}{2}(x_i+t_i+1)^3.
\]

**Start:** $x_i^{(0)}=t_i(t_i-1)$

## 29. Discrete Integral Equation Function {#problem-29-disc-ie}

**Dimensions:** $n\ge1$, $m=n$; default $n=35$<br>
**Factory:** [`disc_ie()`](../reference/disc_ie.html)

Let $h=1/(n+1)$, $t_i=ih$, and $q_j(x)=(x_j+t_j+1)^3$. For
$i=1,\ldots,n$,

\[
r_i(x)=x_i+\frac{h}{2}\left\{
(1-t_i)\sum_{j=1}^{i}t_jq_j(x)
+t_i\sum_{j=i+1}^{n}(1-t_j)q_j(x)
\right\}.
\]

**Start:** $x_i^{(0)}=t_i(t_i-1)$

## 30. Broyden Tridiagonal Function {#problem-30-broyden-tri}

**Dimensions:** $n\ge1$, $m=n$; default $n=40$<br>
**Factory:** [`broyden_tri()`](../reference/broyden_tri.html)

Set $x_0=x_{n+1}=0$. For $i=1,\ldots,n$,

\[
r_i(x)=(3-2x_i)x_i-x_{i-1}-2x_{i+1}+1.
\]

**Start:** $x^{(0)}=(-1,\ldots,-1)$

## 31. Broyden Banded Function {#problem-31-broyden-band}

**Dimensions:** $n\ge1$, $m=n$; default $n=40$<br>
**Factory:** [`broyden_band()`](../reference/broyden_band.html)

For each $i=1,\ldots,n$, define

\[
B_i=\{j:\max(1,i-5)\le j\le\min(n,i+1),\ j\ne i\}.
\]

Then

\[
r_i(x)=x_i(2+5x_i^2)+1-\sum_{j\in B_i}x_j(1+x_j).
\]

**Start:** $x^{(0)}=(-1,\ldots,-1)$

## 32. Linear Function - Full Rank {#problem-32-linfun-fr}

**Dimensions:** $n\ge1$, $m\ge n$; defaults $n=45$, $m=100$<br>
**Factory:** [`linfun_fr()`](../reference/linfun_fr.html)

Define $c(x)=1+(2/m)\sum_{j=1}^{n}x_j$. Then

\[
r_i(x)=
\begin{cases}
x_i-c(x),&i=1,\ldots,n,\\
-c(x),&i=n+1,\ldots,m.
\end{cases}
\]

**Start:** $x^{(0)}=(1,\ldots,1)$

## 33. Linear Function - Rank 1 {#problem-33-linfun-r1}

**Dimensions:** $n\ge1$, $m\ge n$; defaults $n=45$, $m=100$<br>
**Factory:** [`linfun_r1()`](../reference/linfun_r1.html)

For $i=1,\ldots,m$,

\[
r_i(x)=i\sum_{j=1}^{n}jx_j-1.
\]

**Start:** $x^{(0)}=(1,\ldots,1)$

## 34. Linear Function - Rank 1 with Zero Columns and Rows {#problem-34-linfun-r1z}

**Dimensions:** $n\ge1$, $m\ge n$; defaults $n=45$, $m=100$<br>
**Factory:** [`linfun_r1z()`](../reference/linfun_r1z.html)

Define $s(x)=\sum_{j=2}^{n-1}jx_j$. The residuals are

\[
\begin{aligned}
r_1(x)&=-1,\\
r_i(x)&=(i-1)s(x)-1,&&i=2,\ldots,m-1,\\
r_m(x)&=-1.
\end{aligned}
\]

**Edge case:** For $n\le2$, the sum defining $s(x)$ is empty.

**Start:** $x^{(0)}=(1,\ldots,1)$

## 35. Chebyquad Function {#problem-35-chebyquad}

**Dimensions:** $n\ge1$, $m=n$; default $n=50$<br>
**Factory:** [`chebyquad()`](../reference/chebyquad.html)

Let $T_i^*(u)=T_i(2u-1)$ be the Chebyshev polynomial shifted to $[0,1]$,
with

\[
\begin{aligned}
T_0^*(u)&=1,\\
T_1^*(u)&=2u-1,\\
T_i^*(u)&=2(2u-1)T_{i-1}^*(u)-T_{i-2}^*(u).
\end{aligned}
\]

For $i=1,\ldots,n$,

\[
r_i(x)=\frac1n\sum_{j=1}^{n}T_i^*(x_j)-c_i,
\qquad
c_i=
\begin{cases}
0,&i\text{ odd},\\
-1/(i^2-1),&i\text{ even}.
\end{cases}
\]

**Package note:** MGH defines Chebyquad for $m\ge n$, with residual indices
through $m$. The current package has no independent `m` control and implements
only $m=n$; the narrower package domain is the supported contract here.

**Start:** $x_j^{(0)}=j/(n+1)$
